---
title: "Drawing the Borders of Olfactory Space"
author: Emily J Mayhew, Charles J Arayata, Richard C. Gerkin, Brian K. Lee, Jonathan Magill, Lindsey L Snyder, Kelsie Little, Chung
  Wen Yu, Joel D Mainland
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    df_print: paged
    fig_height: 4.5
    fig_width: 7
    highlight: tango
    number_sections: yes
    theme: cosmo
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# In this markdown

This markdown trains models, conducts analyses, and generates figures presented in the manuscript Drawing the Borders of Olfactory Space (bioRxiv preprint: https://doi.org/10.1101/2020.12.04.412254).

Complete datasets will be uploaded at a later date.


```{r, echo = T}
#load required packages 
pacman::p_load(tidyverse, caret, FactoMineR, DMwR, nnet, ROCR, plotly, knitr, RRF, ggmap, cowplot)

```

# Process data

## Read in data tables

These data tables will be made available at a later date.

```{r}
#read in supplemental data tables
S1 <- read.csv("tables/S1.csv", stringsAsFactors = F)
S2 <- read.csv("tables/S2.csv", stringsAsFactors = F)
S3 <- read.csv("tables/S3.alt.csv", stringsAsFactors = F)

```


## Partition data

The dataset was divided into 3 parts: a training set (n=1869), a validation set for model tuning (n=30), and a test set for measuring final model performance (n=30).

```{r}
#make odor class a factor
S1 <- S1 %>%
  mutate(odor.class = factor(odor.class,
                              levels = c("Odorless", "Odor")))

#split dataset based on assignment in "dataset.assigned" column
train <- S1 %>% filter(dataset.assigned != "test")
#validation <- S1 %>% filter(dataset.assigned == "validation") #combined with train for final model
test <- S1 %>% filter(dataset.assigned == "test") 


```


## Pre-process and rebalance training set

```{r, warning=F}
#select only columns needed for model training
train <- train %>% select(odor.class, boiling.point.estimate.Banks:Infective.50)

#generate pre-processing procedure to scale (0 to 1), drop variables with near zero variance, drop highly correlated variables
set.seed(111)
preProc <- preProcess(train[,-1], method = c("range", "zv", "corr"), na.remove = TRUE, verbose = FALSE, k=5, freqCut = 95/5, uniqueCut = 10, cutoff = 0.99, rangeBounds = c(0, 1))

#apply pre-processing to datasets
train_scaled <- predict(preProc, train) 
test_scaled <- predict(preProc, test)
ALL_scaled <- predict(preProc, S1)

#apply SMOTE procedure to rebalance odor:odorless examples
set.seed(123)
#uses scaled data (for everything model)
train_smote <- SMOTE(odor.class ~., train_scaled, perc.over = (100 + 50*6), k=5, perc.under = 200)  

```


# Train models

```{r}
# set seeds amd training parameters

#generate list of seeds to use in training models
set.seed(724)
train.seeds <- sample(1:999, 10, replace=FALSE)

#set training parameters
fit.control <- trainControl(
  method = "repeatedcv", 
  number = 5, 
  repeats = 2,
  savePredictions = T,
  returnResamp = "all",
  classProbs = T,
  summaryFunction = twoClassSummary)

#regularized xgb-linear tuning grid
xgbTuningGrid = expand.grid(nrounds = c(50, 100), 
                            lambda = seq(0.1, 0.5, 0.1), 
                            alpha = seq(0.1, 0.5, 0.1),
                            eta = c(0.3, 0.4))

```


## Train transport ML model

```{r}
### ML Transport model ###

if (file.exists("xgbL.transport.rds")){
  xgbL.transport <- readRDS("xgbL.transport.rds")
}else {
  
#subset train dataset to only transport features
train_transport <- train_smote[,c("odor.class", "boiling.point.best.available", "vapor.pressure.best.available", "MLOGP")]

#train model
set.seed(train.seeds[6])
      xgbL.transport <- train(odor.class~., data=train_transport, 
                         method = "xgbLinear", 
                         trControl = fit.control,
                         metric = "ROC", 
                         tuneGrid = xgbTuningGrid, 
                         na.action = na.roughfix) 
saveRDS(xgbL.transport, "xgbL.transport.rds")
}
```


## Train many-feature ML model

```{r}
### ML Many Feature model ###

if (file.exists("xgbL.manyfeature.rds")){
  xgbL.manyfeature <- readRDS("xgbL.manyfeature.rds")
}else {
#subset by dropping estimated boiling point values, keep only "best available" boiling point
train_manyFeature <- train_smote %>% select(odor.class, boiling.point.best.available:Infective.50)

#train model      
set.seed(train.seeds[6])
      xgbL.manyfeature <- train(odor.class~., data=train_manyFeature, 
                         method = "xgbLinear", 
                         trControl = fit.control,
                         metric = "ROC", 
                         tuneGrid = xgbTuningGrid, 
                         na.action = na.roughfix) 
saveRDS(xgbL.manyfeature, "xgbL.manyfeature.rds")
}
```


# Reproduce Figures

## Figure 1

Fig. 1. A machine learning model can accurately classify molecules as odorous or odorless based only on transport features. 

```{r, results='hide'}
#######################################-
#Figure 1 Associated Stats ------
#######################################-

#dataset composition
table(S1$odor.class) # 1615 odor, 309 odorless
nrow(S1[S1$odor.class == "Odor",])/nrow(S1) # 83.9% odorous
nrow(S1[S1$odor.class == "Odorless",])/nrow(S1) # 16.1% odorless

```


### Figure 1B

Transport-feature ML model-generated odorous probabilities for all molecules in the dataset. Each dot represents one molecule colored by the ground truth, and the width of the violin plot is the density of molecules at a given prediction value. 

```{r}
#######################################-
# Figure 1B: Transport ML Model predictions on full dataset
#######################################-

#generate predictions for transport model on full dataset 
prob.transport <- predict(xgbL.transport, newdata = ALL_scaled, type = "prob")[,2]

#combine predictions with true odor class
predictions <- ALL_scaled %>% 
  select(odor.class) %>%
  mutate(p.transport = prob.transport)
  
#Make dotplot + violin of model predictions
set.seed(123)
ggplot(predictions, aes(odor.class, p.transport, color = odor.class)) +
  geom_jitter(alpha=0.25) +
  geom_violin(color = "black", alpha = 0.1, scale = "width")+
  coord_flip()+
  theme_classic()+
  labs(x = "True Odor Classification",
       y = "Transport ML Model Predicted Odorous Probability")+
  scale_color_brewer(type = "qual", palette = 6, direction = -1)+
  theme(axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8),
        axis.title.x = element_text(size = 9),
        axis.title.y = element_text(size = 9),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position="none")

#ggsave("figures/final/Fig1B.png", width = 3.6, height = 2)
#ggsave("figures/final/Fig1B.pdf", width = 4, height = 1.8)
```


Model performance on a held-out test set of 30 molecules, quantified by the metric Area Under the Receiver Operating Characteristic (AUROC) curve, for transport-feature and many-feature machine learning models. 

```{r, results='hide', warning=F}
#######################################-
# Model performance: Transport ML vs All feature ML and Literature models AUROC ----
#######################################-

#load function to calculate AUROC SD
source("code/AUROCSDfn.R")

#print class ratios - needed in calculation of AUROC SE
table(test$odor.class) #20 odor/10 odorless


#Many feature ML performance on validation set ----

#generate ML predictions on validation set molecules
ML.test.prob <- predict(xgbL.manyfeature, newdata = test_scaled, type = "prob")[,1]
#ML.prob <- predict(xgbL.manyfeature, newdata = ALL_scaled, type = "prob")[,1]
#saveRDS(ML.prob, "Data/Processed/ML.prob.rds") #for supp tables

#compare many feature ML odor class probabilities to true class
ML.test.pred <- prediction(ML.test.prob, test$odor.class)
#calculate auroc
ML.test.perf <- performance(ML.test.pred, "auc") 
#print auroc
ML.test.perf@y.values[[1]] #AUROC = 0.94

#generate raw odor/odorless classifications (instead of probabilities)
ML.test.raw <- predict(xgbL.manyfeature, newdata = test_scaled, type = "raw")
#generate confusion matrix to extract # misclassified
confusionMatrix(ML.test.raw, test_scaled$odor.class)
#calculate SE - inputs are # minor class, # major class, # misclassified
ML.SE <- AUROC.CI(10, 20, 6) #0.0757


#Transport feature ML performance on validation set ----

#generate ML transport predictions on validation set molecules
test.prob <- predict(xgbL.transport, newdata = test_scaled, type = "prob")[,1] 
#compare transport ML odor class probabilities to true class
test.pred <- prediction(test.prob, test_scaled$odor.class)
#calculate auroc
test.perf <- performance(test.pred, "auc") 
#print auroc
test.perf@y.values[[1]] #AUROC = 1.0

#generate raw odor/odorless classifications (instead of probabilities)
test.raw <- predict(xgbL.transport, newdata = test_scaled, type = "raw")
#generate confusion matrix to extract # misclassified
confusionMatrix(test.raw, test_scaled$odor.class)
#calculate SE - inputs are # minor class, # major class, # misclassified
tML.SE <- AUROC.CI(10, 20, 2) #0.0410


#Validation set metrics ----
AUROCcomp <- data.frame(Model = c("Transport ML", "Many-Feature ML"),
                        AUROC = c(test.perf@y.values[[1]], ML.test.perf@y.values[[1]]),
                        SE = c(tML.SE, ML.SE)) %>%
  mutate(SE.upper = ifelse(AUROC + SE < 1, SE, 1 - AUROC))
```

```{r, warning=F}
#######################################-
# Model performance: Transport ML vs All feature ML and Literature models AUROC ----
#######################################-
paste("The transport model AUROC on the test set is", test.perf@y.values[[1]] , "±", round(tML.SE,2))

paste("The many feature model AUROC on the test set is", ML.test.perf@y.values[[1]] , "±", round(ML.SE,2))
```

### Figure 1C

Density of odorous and odorless molecules in transport space defined by molecular weight and number of heteroatoms. Each successive contour line indicates a step increase in density (red = 0.05%, blue = 0.015%). Molecules have a discrete number of heteroatoms, but are jittered along the y-axis to better show density. Plotted within the black box, molecules that obey the “rule of three” with fewer than 3 heteroatoms and molecular weight between 30 and 300 g/mol are generally odorous. 

```{r, results='hide'}
#######################################-
#Figure 1C: Rule of Three figure ------
#######################################-

#print ratio of odorous:odorless molecules in dataset
table(S1$odor.class) #1615 odor, 309 odorless

#finding the best easy filter
S1 %>%
  filter(MW > 30 & MW < 300) %>%
  filter(nHet <3)%>%
  select(odor.class) %>%
  table()

#RULE OF 3: 30 < MW < 300 & nHet <3
#1094 odor, 49 odorless
1094/(1094+49) #95.7% odors
1094/1615 #captures 67.7% of odors = sensitivity
(309-49)/309 #excludes 84.1% odorless molecules = specificity
(1094/(1615-1094))/(49/(309-49)) #odds ratio = odds pass filter is odor/odds pass filter if not = 11.1
```

```{r, warning=F}
#density map
set.seed(123)
ggplot(S1) + 
  geom_jitter( aes(x=MW, y=nHet, color=odor.class), alpha = 0.1, size=2)+
  scale_colour_brewer(type="qual", palette = 6, direction = -1) +
  geom_density_2d(data = S1[S1$odor.class == "Odor",], aes(x=MW, y=nHet), color = "#e41a1c") + 
  geom_density_2d(data = S1[S1$odor.class == "Odorless",], aes(x=MW, y=nHet), color = "#377eb8") + 
  xlab("Molecular Weight (g/mol)") + ylab("Number of Heteratoms") + 
  scale_x_continuous(limits = c(0,500),expand = c(0.01, 0.01), breaks = c(0, 30, 100, 200, 300, 400, 500))+
  scale_y_continuous(limits = c(-0.5,12),expand = c(0.01, 0.01))+
  geom_segment(aes(x=30, y=2.5, xend=300, yend=2.5), color = "black", size=0.3)+
  geom_segment(aes(x=30, y=-0.5, xend=30, yend=2.5), color = "black", size=0.3)+
  geom_segment(aes(x=300, y=-0.5, xend=300, yend=2.5), color = "black", size=0.3)+
  geom_segment(aes(x=30, y=-0.5, xend=300, yend=-0.5), color = "black", size=0.3)+
  #geom_text( aes(x = 180, y = -0.1, label = "96% odorous"), color = "black", size = 3)+
  labs(color = "Odor Class")  +
  theme_classic() +
  theme(axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8),
        axis.title.x = element_text(size = 9),
        axis.title.y = element_text(size = 9),
        legend.text = element_text(size=8),
        legend.title = element_text(size=9),
        #legend.position = c(0.2, 0.9))
        legend.position = "right")

#ggsave("figures/final/Fig1D.png", height = 4.2, width = 4.2)
#ggsave("figures/final/Fig1D.pdf", height = 2.5, width = 4)
```

```{r, results='hide'}
#extract contour densities for Odorless compounds
gg <- ggplot(S1) + 
  geom_density_2d(data = S1[S1$odor.class == "Odorless",], aes(x=MW, y=nHet) ) 
gb <- ggplot_build(gg)
gbd <- gb$data[[1]]
unique(gbd$level) #1e-04 2e-04 3e-04 4e-04 5e-04 6e-04 7e-04 


#extract contour densities for Odorous compounds
gg2 <- ggplot(S1) + 
    geom_density_2d(data = S1[S1$odor.class == "Odor",], aes(x=MW, y=nHet)) 
gb <- ggplot_build(gg2) 
gbd <- gb$data[[1]]
unique(gbd$level) # = 0.0005 0.0010 0.0015 0.0020 0.0025

```


## Figure 2

Fig. 2. Inaccuracies in data impact model performance. 

### Figure 2A

Difference between experimentally determined boiling point (BP) values and BP values calculated using the Burnop (8) and Banks (9) methods. 

```{r, warning=F}
#######################################-
#Figure 2A: Compare BP error of Banks and Burnop eqns-----
#######################################-

#plot error in Burnop and Banks estimates of boiling point compared with experimental values
S1 %>%
  filter(nC + nO + nN + nF + nCL + nBR + nI == nSK) %>% #filter to molecules w/i domain of Burnop applicability
  filter(!is.na(boiling.point.experimental))%>%
  mutate(Banks = (boiling.point.estimate.Banks - boiling.point.experimental))%>%
  mutate(Burnop = (boiling.point.estimate.Burnop - boiling.point.experimental))%>%
  pivot_longer(Banks:Burnop, names_to = "Method", values_to = "BPerror")%>%
  ggplot( aes(BPerror, color = Method))+
  geom_density()+
  xlim(-750,750)+
  xlab("Boiling Point Error in °C\n(Estimated - Experimental)")+
  ylab("Relative frequency")+
  theme_classic()+
  scale_color_brewer(type="qual", palette = 3)+
  theme(axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8),
        axis.title.x = element_text(size = 9),
        axis.title.y = element_text(size = 9),
        legend.title = element_text(size=9))

#ggsave("figures/final/Fig2A.png", width = 3.5, height = 3)
#ggsave("figures/final/Fig2A.pdf", width = 4, height = 2)

```


### Figure 2B

Odor classification predictions by transport-feature ML models using only estimated BP values calculated by the  Burnop method. 

```{r}

#######################################-
#Figure 2B: Burnop BP Estimate model accuracy-----
#######################################-

if (file.exists("xgbL.Burnop.rds")){
  xgbL.Burnop <- readRDS("xgbL.Burnop.rds")
}else {
#filter training set
train_Burnop <- train_smote %>% select(odor.class, boiling.point.estimate.Burnop, MLOGP) %>%
  filter(! is.na(boiling.point.estimate.Burnop))

#train model
set.seed(train.seeds[6])
      xgbL.Burnop <- train(odor.class~., data=train_Burnop, 
                         method = "xgbLinear", 
                         trControl = fit.control,
                         metric = "ROC", 
                         tuneGrid = xgbTuningGrid, 
                         na.action = na.roughfix) 
#save model
saveRDS(xgbL.Burnop, "xgbL.Burnop.rds")  
}
```

```{r, results='hide'}

#view max cross-validation AUROC
max(xgbL.Burnop$results$ROC) #CV AUROC 0.93

#Burnop BP estimate model predictions
Burnop.prob <- predict(xgbL.Burnop, newdata = ALL_scaled, type = "prob")[,2] 

#combine predictions with true odor class
Burnop_predictions <- ALL_scaled %>% 
  filter(! is.na(boiling.point.estimate.Burnop)) %>%
  select(odor.class) %>%
  mutate(p.transport = Burnop.prob)
```

```{r}
#Make dotplot + violin of model predictions - Burnop edition
set.seed(123)
ggplot(Burnop_predictions, aes(odor.class, p.transport, color = odor.class)) +
  geom_jitter(alpha=0.25) +
  geom_violin(color = "black", alpha = 0.1, scale = "width")+
  coord_flip()+
  theme_classic()+
  labs(x = "True Odor Classification",
       y = "Predicted Odorous Probability using Burnop-Estimated BP")+
  scale_color_brewer(type = "qual", palette = 6, direction = -1)+
  theme(axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8),
        axis.title.x = element_text(size = 9),
        axis.title.y = element_text(size = 9),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position="none")

#ggsave("figures/final/Fig2B.png", width = 4, height = 2)
#ggsave("figures/final/Fig2B.pdf", width = 4, height = 1.8)
```


### Figure 2C

Odor classification predictions by transport-feature ML models using only estimated BP values calculated by the Banks method.

```{r}
#######################################-
#Figure 2C: Banks BP Estimate model accuracy-----
#######################################-

if (file.exists("xgbL.Banks.rds")){
  xgbL.Banks <- readRDS("xgbL.Banks.rds")
}else {
#filter training set
train_Banks <- train_smote %>% select(odor.class, boiling.point.estimate.Banks, MLOGP) %>%
  filter(! is.na(boiling.point.estimate.Banks))


S1 %>% filter(is.na(boiling.point.estimate.Banks)) %>% nrow()

#train model
set.seed(train.seeds[6])
      xgbL.Banks <- train(odor.class~., data=train_Banks, 
                         method = "xgbLinear", 
                         trControl = fit.control,
                         metric = "ROC", 
                         tuneGrid = xgbTuningGrid, 
                         na.action = na.roughfix) 
#save model
saveRDS(xgbL.Banks, "xgbL.Banks.rds")
}
```

```{r, results='hide'}

#view max cross-validation AUROC
max(xgbL.Banks$results$ROC)  #CV AUROC 0.95

#Banks BP estimate model predictions
Banks.prob <- predict(xgbL.Banks, newdata = ALL_scaled, type = "prob")[,2] 

#combine predictions with true odor class
Banks_predictions <- ALL_scaled %>% 
  filter(! is.na(boiling.point.estimate.Banks)) %>%
  select(odor.class) %>%
  mutate(p.transport = Banks.prob)
```

```{r}
#Make dotplot + violin of model predictions
set.seed(123)
ggplot(Banks_predictions, aes(odor.class, p.transport, color = odor.class)) +
  geom_jitter(alpha=0.25) +
  geom_violin(color = "black", alpha = 0.1, scale = "width")+
  coord_flip()+
  theme_classic()+
  labs(x = "True Odor Classification",
       y = "Predicted Odorous Probability using Banks-Estimated BP")+
  scale_color_brewer(type = "qual", palette = 6, direction = -1)+
  theme(axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8),
        axis.title.x = element_text(size = 9),
        axis.title.y = element_text(size = 9),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position="none")

#ggsave("figures/final/Fig2C.png", width =4, height = 2)
#ggsave("figures/final/Fig2C.pdf", width =4, height = 1.8)
```


### Figure 2D

Human subject-classified molecules in transport space defined by BP and log P. Many clearly non-volatile molecules were initially classified as odors due to odorous contaminants.

```{r, results='hide'}
#######################################-
#Figure 2D: Reclassified molecules in transport feature space-----
#######################################-

#what percent of lab-tested molecules initially classified as Odorous were reclassified through QC process?
nrow(S2[S2$initial.odor.class == "Odor",]) #111 Odors initially
nrow(S2[S2$QC.reclassified == "yes",]) #24 reclassified
nrow(S2[S2$QC.reclassified == "yes",])*100/nrow(S2[S2$initial.odor.class == "Odor",]) #21.6% reclassified

#create small df with lab-tested molecules, reclassification info - PLOT 3 CATEGORIES
reclass <- S1 %>%
  filter(odor.class.source == "Lab-Tested") %>%
  left_join(S2, by = "SMILES") %>%
  select(compound.name, odor.class, QC.reclassified, boiling.point.best.available, MLOGP, nHet) %>%
  mutate(odor.class = as.character(odor.class))%>%
  mutate(odor.3class = ifelse(QC.reclassified == "yes", "Odorless with odorous contaminants", odor.class)) %>%
  mutate(BP.K = boiling.point.best.available + 273.15) 
```

```{r}
#plot molecules, showing reclassified molecules, in transport space 
set.seed(123)
ggplot(reclass, aes(x=boiling.point.best.available, y=MLOGP, color=odor.3class, shape=odor.3class)) + 
  geom_jitter(alpha = 0.75) +
  theme_classic() +
  scale_color_manual(values = c("#e41a1c", "#377eb8", "black"))+
  scale_shape_manual(values = c(16, 16, 4))+
  #scale_x_log10()+
  theme(legend.position = "top",
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8),
        axis.title.x = element_text(size = 9), 
        axis.title.y = element_text(size = 9),
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 9),
        legend.key.width = unit(0.05, "in")) +
  xlab("Boiling Point (°C)") + 
  #ylab("Number of Heteratoms") + 
  ylab("Log P") + 
  labs(color = "Class", shape = "Class")   


#ggsave("figures/final/Fig2D.png", height = 3, width = 4)
#ggsave("figures/final/Fig2D.pdf", height = 3, width = 4)
```


### Figure 2E

Transport-feature ML model odor predictions for human subject-classified molecules. Chemical compounds that are odorless but had odorous contaminants are correctly predicted to be odorless by the model. 
 
```{r}
#######################################-
#Figure 2E: Validation set predictions - models catch misclassifications-----
#######################################-

#generate predictions for transport ML model on full dataset 
prob.transport <- predict(xgbL.transport, newdata = ALL_scaled, type = "prob")[,2]

#combine predictions with descriptive data
LabTest_predictions <- ALL_scaled %>% 
  mutate(p.transport = prob.transport) %>%
  right_join(S2, by = "SMILES") %>%
  select(1:3, QC.reclassified, p.transport) %>%
  mutate(odor.class = as.character(odor.class))%>%
  mutate(odor.3class = ifelse(QC.reclassified == "yes", "Odorless with odorous contaminants", odor.class))
  
#Make dotplot of model predictions, showing reclassified molecules
set.seed(123)
ggplot(LabTest_predictions, aes(odor.3class, p.transport, color = odor.3class, shape = odor.3class)) +
  geom_jitter(alpha=0.75) +
  coord_flip()+
  theme_classic()+
  labs(x = "True Odor Classification",
       y = "Transport ML Model Predicted Odorous Probability")+
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10))+
  scale_color_manual(values = c("#e41a1c", "#377eb8", "black"))+
  scale_shape_manual(values = c(16, 16, 4))+
  theme(axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8),
        axis.title.x = element_text(size = 9),
        axis.title.y = element_text(size = 9),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position="none")

#ggsave("figures/final/Fig2E.png", width = 4, height = 2.6)
#ggsave("figures/final/Fig2E.pdf", width = 4, height = 2.6)
```

```{r, results='hide'}
#######################################-
#Figure 2 Associated Stats ------
#######################################-

#compare transport ML CV AUROC
max(xgbL.transport$results$ROC) # CV AUROC 0.99

#summary stats of boiling point error
S1 %>%
  filter(!is.na(boiling.point.experimental))%>%
  mutate(Banks = abs(boiling.point.experimental - boiling.point.estimate.Banks))%>%
  mutate(Burnop = abs(boiling.point.experimental - boiling.point.estimate.Burnop))%>%
  pivot_longer(Banks:Burnop, names_to = "Method", values_to = "BPerror")%>%
  group_by(Method)%>%
  summarize(med = median(BPerror, na.rm = T), 
            max = max(BPerror, na.rm = T),
            min = min(BPerror, na.rm = T))

#error rates  of models trained/tested on estimated bp
#banks boiling point error rate
Banks_predictions %>%
  filter(odor.class == "Odor" & p.transport < 0.5 |
           odor.class == "Odorless" & p.transport > 0.5) %>%
  nrow() #114 errors 
114/nrow(Banks_predictions) # 6.0% error rate

#burnop boiling point error rate
Burnop_predictions %>%
  filter(odor.class == "Odor" & p.transport < 0.5 |
           odor.class == "Odorless" & p.transport > 0.5) %>%
   nrow() #86 errors 
86/nrow(predictions) # 4.5% error rate

#compare to error rate using best available boiling point value
predictions %>%
  filter(odor.class == "Odor" & p.transport < 0.5 |
           odor.class == "Odorless" & p.transport > 0.5) %>%
  nrow() #48 errors 
50/nrow(predictions) # 2.6% error rate

### AUROC on original classifications (pre-GC QC)

#add original odor class to test set
test_scaled_initialClass <- test_scaled %>%
  left_join(S2, by = "SMILES") %>%
  select(initial.odor.class, boiling.point.best.available, vapor.pressure.best.available, MLOGP)

#generate ML transport predictions on test set molecules
test.prob.i <- predict(xgbL.transport, newdata = test_scaled_initialClass, type = "prob")[,1] 
#compare transport ML odor class probabilities to true class
test.pred.i <- prediction(test.prob.i, test_scaled_initialClass$initial.odor.class)
#calculate auroc
test.perf.i <- performance(test.pred.i, "auc") 
#print auroc
test.perf.i@y.values[[1]] #AUROC = 0.81

```


## Figure 3

Fig. 3. The transport model can be used to predict the population of odor space and identify its understudied regions.

### Figure 3A

Proportion of molecules predicted by transport ML model to be odorous as a function of heavy atom count (HAC). Red circles show the mean probability generated for HAC tranches from the GDB database (12) with standard error bars; hollow circles show the projected odorous probability generated by the logistic regression fit plotted in black. 

```{r}
### Apply transport model to generate predictions for GDB molecules

#apply model to generate probabilities - using Banks estimate because no experimental values are available
S3$odor.probability <- predict(xgbL.Banks, newdata = S3, type = "prob")[,2]

#number of possible molecules in GDB-17 per HAC
GDB.count <- c(3,6,14,47,219,1091,6029,37435,243233,1670163,12219460,72051665,836687200,
                  2921398415,15084103347,38033661355,109481780580)

#how many odorous molecules? sum odorous probabilities
GDBprob.byHAC <- S3 %>%
  group_by(HAC)%>%
  summarize(meanProb = mean(odor.probability)) %>%
  mutate(possible.molecules = GDB.count) %>%
  mutate(Odors = meanProb*possible.molecules)

```

```{r, results='hide'}
#generate random binarizations from predicted odorous probabilities
set.seed(111)
for(i in 1:nrow(S3)){
  S3$Trial1[i] <- rbinom(1,1,S3$odor.probability[i])
  S3$Trial2[i] <- rbinom(1,1,S3$odor.probability[i])
  S3$Trial3[i] <- rbinom(1,1,S3$odor.probability[i])
  S3$Trial4[i] <- rbinom(1,1,S3$odor.probability[i])
  S3$Trial5[i] <- rbinom(1,1,S3$odor.probability[i])
  S3$Trial6[i] <- rbinom(1,1,S3$odor.probability[i])
  S3$Trial7[i] <- rbinom(1,1,S3$odor.probability[i])
  S3$Trial8[i] <- rbinom(1,1,S3$odor.probability[i])
  S3$Trial9[i] <- rbinom(1,1,S3$odor.probability[i])
  S3$Trial10[i] <- rbinom(1,1,S3$odor.probability[i])
}
#how variable are the trials?
S3 %>%
  select(Trial1:Trial10)%>%
  colSums() #not very variable

#use data from all 10 simulations
S3.long <- S3 %>%
  pivot_longer(Trial1:Trial10, names_to = "TrialNo", values_to = "Class")

#apply glm-binomial to binomial data for 10 simulations
t1 <- glm(Class ~ HAC, data = S3.long, family = "binomial")
summary(t1)
eq.p <- function (x) {exp(-0.3361550*x + 4.102855)/(1 + exp(-0.336155*x + 4.102855))} #need to manually update from summary(t1)!

#test predicted % odorous for HAC 18-21 -> 4.5% for HAC 21
eq.p(21)

#extend to HAC 21
HAC21 <- tibble(HAC = 18:21,
                Prob.Odor = eq.p(18:21))

#calculate variance
S3 <- S3 %>%
  mutate(Variance = odor.probability*(1-odor.probability))
sum(S3$Variance) #12,825.17
```


```{r, warning=F}
#######################################-
#Figure 3A: Enumerating odor space-Percent Odorous -----
#######################################-

### Make odor probability by HAC plot
plot1 <- S3 %>%
  group_by(HAC) %>%
  mutate(count = n()) %>%
  summarize(Prob.Odor = mean(odor.probability),
            SE = sqrt(sum(Variance))/mean(count))  %>%
  #summarize(Prob.Odor = mean(Class)) %>%
  ggplot( aes(HAC, Prob.Odor)) + 
  geom_point(color="#e41a1c", size=2) +
  geom_point(data=HAC21, aes(HAC, Prob.Odor), shape=21, size=2)+
  geom_errorbar( aes(ymin = Prob.Odor - SE, ymax = Prob.Odor+ SE))+
  #geom_violin(aes(group=heavy.atoms)) +
  #geom_bar(stat = "identity", position = "stack")  + 
  stat_function(fun=eq.p, xlim = c(3,21)) +
  #geom_smooth()+
  #scale_x_continuous(limits = c(0,21), expand = c(0.01, 0.01), breaks = seq(1,21,4)) +
  scale_x_continuous(limits = c(0.5,21.5), expand = c(0.05, 0.01), breaks = seq(1,21,1)) +
  scale_y_continuous(limits = c(0,1.0), expand = c(0.02, 0), breaks = seq(0,1.00,0.10)) +
  #theme_minimal() + 
  theme_classic()+
  xlab("Number of Heavy Atoms") + 
  ylab("Proportion Molecules \n Predicted to be Odorous") +
  labs(fill = "Prediction") + 
  theme(plot.title = element_text(hjust = 0.5), 
        text = element_text(size=8),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank())
plot1
```


### Figure 3B

Estimated number of possible molecules and predicted odorous molecules from the GDB databases as a function of HAC. Lighter shaded bars indicate HAC values beyond GDB-17; as a conservative estimate, the number of possible molecules with HAC 18-21 was set to equal that of HAC 17. 

```{r}
#######################################-
#Figure 3B: Enumerating odor space-Number predicted molecules & odors -----
#######################################-

#make summary table with predicted # odorous molecules by HAC, use HAC-17 molecule count for HAC 18-21
GDBTotal.Extrapolate <- tibble(HAC = 1:21,
                               Source = c(rep("GDB", 17), rep("Curve Fit", 4)),
                               Molecules = c(3,6,14,47,219,1091,6029,37435,243233,1670163,12219460,72051665,836687200,
                                             2921398415,15084103347,38033661355, rep(109481780580, 5)),
                               Prob.Odor = eq.p(1:21),
                               Predicted.Odor = Molecules*Prob.Odor)

#format df to generate plot
p <- GDBTotal.Extrapolate %>% gather(key = "type", value = "number", -c(1:2,4)) %>% 
  mutate(number = number/1E9)%>%
  mutate(type = factor(type, levels = c("Molecules", "Predicted.Odor"))) 
p$type <- gsub("Predicted.Odor", "Odorous Molecules", p$type)

#plot bar graph of # molecules and predicted # odorous molecules as a function of HAC
plot2 <- ggplot() + 
  geom_bar(data=subset(p, HAC<18), aes(HAC, number, fill=type, color=type), stat = "identity", position = "dodge")  + 
  geom_bar(data=subset(p, HAC>17), aes(HAC, number, fill=type), stat = "identity", position = "dodge", alpha = 0.3)  + 
  scale_x_continuous(limits = c(0.5,21.5), expand = c(0.05, 0.01), breaks = seq(1,21,1)) +
  theme_classic() + 
  scale_fill_brewer(name = "Prediction", 
                    labels = c("Molecules", "Odorous Molecules"),
                    type="qual", palette = 6, direction = -1) + 
  scale_color_brewer(name = "Prediction", 
                     labels = c("Molecules", "Odorous Molecules"),
                     type="qual", palette = 6, direction = -1) + 
  xlab("Number of Heavy Atoms") + 
  ylab("Predicted Molecules \n and Odors (Billions)") +
  #labs(fill = "Prediction") + 
  scale_y_continuous(limits = c(0,120),
                     breaks = c(0, 20, 40, 60, 80, 100, 120),
                     expand = c(0, 0)) +
  theme(plot.title = element_text(hjust = 0.5), 
        text = element_text(size=8),
        legend.position = c(0.2, 0.8), 
        axis.line.x = element_blank())
plot2
```


### Figure 3C

Cumulative estimates of possible molecules and odorous molecules with increasing HAC on a logarithmic scale. The red datapoint at HAC 17 reflects our conservative estimate of 33 billion odorous molecules; the light red datapoint at HAC 21 reflects our less-conservative estimate of 62 billion odorous molecules.

```{r}
#######################################-
#Figure 3C: Enumerating odor space-Cumumulative predicted molecules & odors -----
#######################################-

#calculate cumulative molecules and predicted odorous molecules as a function of HAC 
p2 <- GDBTotal.Extrapolate %>%
  mutate(Cumulative.Molecules = cumsum(Molecules),
         Cumulative.Odors = cumsum(Predicted.Odor)) %>%
  gather(key = "type", value = "number", 6:7) %>%
  mutate(type = factor(type, levels = c("Molecules", "Odorous Molecules"))) 
p2$type <- gsub("Cumulative.Molecules", "Molecules", p$type)
p2$type <- gsub("Cumulative.Odors", "Odorous Molecules", p$type)

#plot cumulative molecules and odorous molecules by HAC on a log scale
plot3 <- ggplot()+
  geom_point(data=subset(p2, HAC<18), aes(HAC, number, color=type), size=2)  + 
  geom_point(data=subset(p2, HAC>17), aes(HAC, number, color=type), size=2, alpha = 0.3)  + 
  scale_x_continuous(limits = c(1,21.5), expand = c(0.05, 0.1), breaks = seq(1,21,1)) +
  theme_classic() + 
  scale_color_brewer(type="qual", palette = 6, direction = -1) + 
  xlab("Number of Heavy Atoms") + 
  ylab("Cumulative Predicted Molecules \n and Odors (log scale)") +
  labs(color = "Prediction") + 
  scale_y_log10(limits=c(1E0, 1E14),
                breaks = scales::trans_breaks("log10", function(x) 10^x),
                labels = scales::trans_format("log10", scales::math_format(10^.x)),
                expand = c(0, 0)) +
  theme(plot.title = element_text(hjust = 0.5), 
        text = element_text(size=8),
        legend.position = c(0.2, 0.8))

plot3
```

```{r}
### print odor space enumeration estimates

#how many odorous molecules? sum odorous probabilities for GDB-17
paste0("A conservative estimate of the number of odorous molecules with 17 or fewer heavy atoms is ", round(sum(GDBprob.byHAC$Odors),0))

#new extrapolated estimate for HAC < 22
paste0("A conservative estimate of the number of odorous molecules with 21 or fewer heavy atoms is ", round(sum(GDBTotal.Extrapolate$Predicted.Odor),0))
```


```{r, eval=F}
#combine figure 3 plots on shared x axis
cowplot::plot_grid(plot1, plot2, plot3, align = "v", ncol = 1, rel_heights = c(0.35, 0.3, 0.35))

#ggsave("figures/final/Fig4ABC.png", height = 6, width = 4)
#ggsave("figures/final/Fig3.pdf", height = 6, width = 4)

```

